---
title: "R Programming for Geospatial Applications (GEOG4254/5254G)"
author: "Yang Shao"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
  word_document:
    toc: true
    toc_depth: 2
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Information on this page has not been updated since `r format(Sys.Date(), "%B %d, %Y")`. Students should go to Canvas for the most up-to-date course syllabus and materials.


# Course Description

Programming is increasingly important in geospatial research and applications. Students in the geospatial field are now required to learn basic programming skills in R, Python, and Matlab, and develop batch-processing capabilities for handling large geospatial data (e.g., time-series remote sensing imagery, various GIS layers). Many geospatial tasks can be automated using these programming languages. Students are often encouraged to learn multiple programming languages or tools to effectively communicate with other professionals.

This course, GEOG4254/5254, focuses on student-centered project work through laboratory sessions and independent project homework. All aspects of a GIS/RS project will be developed using R programming. Throughout the semester, students will work on 7-8 programming projects. They will apply their knowledge of GIS and remote sensing data manipulation to various application areas to develop automated GIS/RS data processing workflows. The course aims to improve students’ technical proficiency in R programming and their ability to solve spatial problems through properly sequenced operations of their own design.

# Learning Outcomes/Goals

- **Proficiency in R Programming:** Students will become proficient in using R to retrieve and analyze geospatial data.
- **Problem-Solving:** Students will be capable of using R to model and solve moderately difficult GIS and remote sensing problems.

# Course Prerequisites
Introduction to Python/R; Geog4084/5084 Modelling with GIS

# Instructor: 
Yang Shao, 217 Wallace Hall, email: yshao@vt.edu

# Lectures and lab sessions: 
Tuesday and Thursday – 9:30am– 10:45am

# Office Hours: 
Wednesday 3:30PM-4:30PM

# Materials: 
Laptop with RStudio and ArcGIS installed (optional) 

# Grading: 
Labs and assignments - 75%
Midterm exam (or project) – 10%
Final project – 15%

Labs and homework assignments are designed not only to give you experience with R programming, but also to have you apply the concepts that we have been discussing in class. You can discuss the lab and get assistance from instructor and labmates, but all answers should be your own.  Ten percent of the grade will be deducted for each day late. 

Grading scale (A = 90 to 100%, B=80 to 89%, C=70 to 79%, D=60 to 69%, F=0 to 59%). 

# How to succeed in this course:
- Attend lectures and labs consistently
- If something is not clear, ask a question
- Clarify expectations with the instructor

# Student Responsibilities: 
It is your responsibility to attend all lectures. Most of the hands-on sessions consist of a walk-through of techniques vital to the successful (and timely) completion of your homework projects, and are allowable to make up only if you notify us ahead of time that you cannot attend a particular week. If you miss a lecture, it is your responsibility to obtain suitable notes from a classmate who attended that day.

"The honor code will be strictly enforced in this class. All assignments submitted shall be considered graded work unless otherwise noted. All aspects of your course work are covered by the honor system. Honesty in your academic work will develop into professional integrity. The faculty and students of Virginia Tech will not tolerate any form of academic dishonesty."

# Services for Students with Disabilities 
Virginia Tech welcomes students with disabilities into the University’s educational programs. The University promotes efforts to provide equal access and a culture of inclusion without altering the essential elements of coursework. If you anticipate or experience academic barriers that may be due to disability, including but not limited to ADHD, chronic or temporary medical conditions, deaf or hard of hearing, learning disability, mental health, or vision impairment, please contact the Services for Students with Disabilities (SSD) office (540-231-3788, ssd@vt.edu, or visit www.ssd.vt.edu). If you have an SSD accommodation letter, please meet with me privately during office hours as early in the semester as possible to deliver your letter and discuss your accommodations. You must give me reasonable notice to implement your accommodations, which is generally 5 business days and 10 business days for final exams.

# Tentative Topics and Schedule

```{r table, echo=FALSE}
suppressMessages({
  library(gt)
  library(dplyr)
})
# Create the schedule data frame
schedule <- data.frame(
  Week = c("Week 1", "Week 2", "Week 3", "Week 4", "Week 5", "Week 6", "Week 7", "Week 8", "Week 9", "Week 10", "Week 11", "Week 12", "Week 13", "Week 14", "Week 15", "Week 16"),
  Dates = c(
    "Aug 25 - Aug 31", "Sep 1 - Sep 7", "Sep 8 - Sep 14", "Sep 15 - Sep 21", "Sep 22 - Sep 28", 
    "Sep 29 - Oct 5", "Oct 6 - Oct 12", "Oct 13 - Oct 19", "Oct 20 - Oct 26", "Oct 27 - Nov 2", 
    "Nov 3 - Nov 9", "Nov 10 - Nov 16", "Nov 17 - Nov 23", "Nov 24 - Nov 30", "Dec 1 - Dec 7", "Dec 8 - Dec 14"
  ),
  Topics = c(
    "Course introduction; R vs. Python vs. Matlab; R Data types; Data frame; Simple plots in R",
    "Read and write geospatial data using R libraries: sf and terra; Projections (Re-project shapefile, re-project raster); Handling spatial data in R (select by attributes, select by location, summary statistics)",
    "Handling spatial data in R (join, buffer, union, Spatial subsetting, raster calculator, zonal statistics; DEM-derivatives)",
     "Create maps in R",
     "R Scripts; Loops and repetition; Conditional Element Selection; Build your R functions",
    "Remote sensing image classification using Random Forest algorithm",
    "Remote sensing image classification; cross-validation; caret package",
    "Regression analysis; Logistic regression",
    "Time-series analysis",
    "Linux environment; R in high-performance computing environment; VT-ARC",
    "Linux environment; R in high-performance computing environment; VT-ARC",
    "Spatial point pattern analysis; Principal components",
    "R for web app; Shiny app",
    "Thanksgiving holiday",
    "Final project (group presentation)",
    "Final project (group presentation)"
  ),
  Labs = c(
    "Install R and RStudio",
    "Lab 1: R fundamentals",
    "Lab 2: Characterizing Wildland–Urban Interface in Virginia (Rscript)",
    "Lab 3: Fire Fuel Mapping and Modeling in a Forested Environment (Rscript)",
    "Lab 4: Working with census data",
    "Lab 5: Image classification with R",
    "Lab 5: Image classification with R",
    "Lab 6: logistic regression for urban growth simulation",
    "Lab 6: logistic regression for urban growth simulation",
    "Lab 7: Getting started with high-performance computing (VT-ARC)",
    "Lab 7: Getting started with high-performance computing (VT-ARC)",
    "Lab 8: Spatial stats",
    "Final project help session",
    "NA",
    "Final project help session",
    "Final project (group presentation)"
  )
)

# Create the table with gt
schedule_table <- gt(schedule) %>%
  tab_header(
    title = "Tentative Topics and Schedule",
    subtitle = "Topics will be adjusted according to participant interest and may be constantly readjusted during the term, along with other activities."
  ) %>%
  cols_label(
    Week = "Week",
    Dates = "Dates",
    Topics = "Course Topics",
    Labs = "Labs/Activities"
  ) %>%
  cols_width(
    c(Week) ~ px(150),
    everything() ~ px(200)
  ) %>%
  tab_options(
    table.font.size = px(12),
    table.width = pct(100)
  )

# Print the table
schedule_table


```

