---
title: "Introduction to R Markdown"
author: "Yang Shao"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction to R Markdown

R Markdown is a powerful tool that allows you to create dynamic documents integrating R code and its output with narrative text. This guide will help you get started with using R Markdown for your assignments.

### Prerequisites

1. **R**: Make sure you have R installed on your computer. You can download it from [CRAN](https://cran.r-project.org/).
2. **RStudio**: Install RStudio, an integrated development environment (IDE) for R, from [RStudio's website](https://www.rstudio.com/products/rstudio/download/).

### Creating an R Markdown Document

1. **Open RStudio**.
2. **Create a New R Markdown File**:
   - Go to `File` -> `New File` -> `R Markdown...`
   - Fill in the details such as the title, author, and default output format (HTML is a good starting point).
   - Click `OK`.

### Structure of an R Markdown Document

An R Markdown document consists of three main parts:

1. **YAML Header**:
   - This section is at the top of the document, enclosed in `---`. It includes metadata like the title, author, and output format.

2. **Markdown Text**:
Use regular Markdown syntax to add narrative text. Markdown allows you to format text using simple symbols.

3. **Code Chunks**:
Code chunks are sections of the document that contain R code. They are enclosed in triple backticks and start with {r}.

```{r}
# This is an R code chunk
summary(cars)
```

Run a Code Chunk:
Click the green triangle (Run Current Chunk) in the chunk header or press Ctrl + Shift + Enter (Cmd + Shift + Enter on Mac) with your cursor inside the chunk.

Adding and Running Code Chunks
Insert a Code Chunk:
You can insert a code chunk by clicking "Insert Chunk" button -> R in the RStudio toolbar. It looks like a small green plus sign (+), or by typing the chunk delimiters manually.

```{r}
plot(cars)
```

### Rendering the Document
Knit the Document:
Click the Knit button at the top of the RStudio script editor. This will generate the output document in the format specified in the YAML header (e.g., HTML).
