---
title: "R Programming for Geospatial Applications"
output:
  powerpoint_presentation: default
  ioslides_presentation: default
date: "2024-07-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The format of this course
- **Lectures and Lab sessions**
- **Office Hours: Wednesday 3:30PM-4:30PM**

- **Grading:**
 
  -Labs and assignments - 75%
 
  -Midterm exam (or project) - 10%
 
  -Final project – 15%

## Outline

- R/Python/Matlab for geospatial research

- Data types and plots in R

- Software preparation (R, Rstudio, sf library, terra library, and mapview)  


## Data types in R
R has a wide variety of data types including scalars, vectors, matrices, data frames, and lists.

Scalar: a single number 
```{r, echo = TRUE}
x <- 1
y <- 3
print(x+y)
```

## Vectors
Using c() or combine function: 
```{r, echo=TRUE}
# numeric vector
x <- c(1,2,5.3,6,-2,4)
# character vector
y <- c('VT','CNRE','Geography')                               
#Refer to elements of a vector using subscripts.
print(y[1:2])
```
## matrices
```{r,, echo=TRUE}
## Define the matrix
m <- matrix(c(1,2,3,3,2,1),nrow=3,ncol=2)
print(m)
## Refer to elements of a matrix:
m[,1]
m[2,]
```
## data frame: 
A data frame is used for storing data tables. It is a list of vectors of equal length. For example, the following variable df is a data frame containing three vectors n, s, b.
```{r,echo=TRUE}
n<-c(2,3,5)
s<-c('aa','bb','cc')
b<-c(TRUE,FALSE,TRUE)
df=data.frame(n,s,b)
print(df)
```

## Lists
A list is an R data structure that can hold elements of different types (like numbers, strings, vectors, and another list). Lists are a powerful and flexible way to group related objects.


```{r,echo=TRUE}
# Create a list
my_list <- list(name = "John", age = 25, scores = c(90, 85, 88))
# Print the list
print(my_list)
# Access elements of a list
print(my_list$name)
print(my_list$scores)
```

## Summary Table
```{r}
library(knitr)
data_types <- data.frame(
  "Data Type" = c("Scalar", "Vector", "Matrix", "Data Frame", "List"),
  "Description" = c("A single number", "A sequence of data elements", "A 2-dimensional array", "A table of data", "A collection of objects"),
  "Example" = c("x <- 1", "x <- c(1, 2, 3)", "m <- matrix(1:6, nrow=2)", "df <- data.frame(x, y, z)", "l <- list(a, b, c)")
)
kable(data_types, format = "markdown")

```

## Key Functions

Here are some key functions for working with data types in R:

*length()*: Get the number of elements in a vector.

*nrow()*, *ncol()*: Get the number of rows and columns in a matrix or data frame.

*str()*: Display the structure of an R object.

*summary()*: Provide a summary of an R object.

*names()*: Get or set the names of elements in a list or data frame.


## Basic R Plots

R provides a variety of plotting functions to create basic visualizations. Here are a few examples:

Scatter Plot - A scatter plot is used to plot points on a graph.
```{r,echo=TRUE}
# Create sample data
x <- rnorm(100)
y <- rnorm(100)

# Create scatter plot
plot(x, y, main="Scatter Plot", xlab="X-axis", ylab="Y-axis", pch=19, col="blue")

```

## Line Plot
A line plot is used to plot lines on a graph.

```{r,echo=TRUE}
# Create sample data
x <- 1:10
y <- c(2, 3, 5, 7, 11, 13, 17, 19, 23, 29)

# Create line plot
plot(x, y, type="o", main="Line Plot", xlab="X-axis", ylab="Y-axis", col="red")

```

## Bar Plot
A bar plot is used to display data with rectangular bars.
```{r,echo=TRUE}
# Create sample data
counts <- table(mtcars$cyl)

# Create bar plot
barplot(counts, main="Bar Plot", xlab="Number of Cylinders", ylab="Frequency", col="green")

```

## Histogram
A histogram is used to display the distribution of a dataset.
```{r,echo=TRUE}
# Create sample data
data <- rnorm(1000)

# Create histogram
hist(data, main="Histogram", xlab="Value", col="purple", border="white")

```

## Box Plot
A box plot is used to display the distribution of data based on a five-number summary.
```{r,echo=TRUE}
# Create sample data
data <- rnorm(100)

# Create box plot
boxplot(data, main="Box Plot", ylab="Values")

```

## Software Preparation: R
Installation
Visit CRAN to download the latest version of R.

https://cran.r-project.org/bin/windows/base/

Follow the installation instructions for your operating system (Windows, macOS, Linux).


## RStudio is an integrated development environment (IDE) for R.
Installation
Visit RStudio to download the latest version of RStudio.

https://posit.co/download/rstudio-desktop/

Install RStudio after installing R.

**Basic Features**

*-Script editor*

*-Console*

*-Environment/History*

*-Plots/Files/Packages/Help*


## Software Preparation

The *sf* package provides simple features for R, making it easy to handle spatial data.

The *terra* package is used for spatial data manipulation, providing functions to create, read, manipulate, and analyze raster and vector data.

The *mapview* package allows for quick and easy interactive visualization of spatial data.

install.packages("sf")

install.packages("terra")

install.packages("mapview")

