---
title: "Lab Instruction: Downloading and Analyzing ACS Data in R"
output: html_document
date: "2024-07-13"
---

## Objective

This lab aims to guide you through downloading American Community Survey (ACS) data, specifically focusing on income at the census tract level. By the end of this lab, you will be able to:

*Access and download ACS data for one or multiple states.*

*Process and clean the data.*

*Perform basic analysis on income data.*

Before starting, make sure you have the following packages installed. 

#install.packages("tidyverse")

#install.packages("tidycensus")

#install.packages("sf")

#install.packages("tigris")

#install.packages("mapview")


## Steps

## Step 1: Load Required Libraries

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(mapview)
```

## Step 2: Set Up Census API Key
You need a Census API key to access ACS data. If you don't have one, you can request it here
at https://api.census.gov/data/key_signup.html Once you have your key, set it up in R.


```{r}
#please replace the API Key
#census_api_key("e23a9a88f3a3911be51aed1a0e9c595a10e35b59", install = TRUE)
census_api_key("e23a9a88f3a3911be51aed1a0e9c595a10e35b59", overwrite = TRUE)
readRenviron("~/.Renviron")
```
## Step 3: Download ACS Data
Use the tidycensus package to download ACS data. We'll focus on median household income and education attainment levels for Virginia.


```{r}
# Define variables for median household income
variables <- c(income = "B19013_001")
# Download data for Virginia
acs_data <- get_acs(
  geography = "tract",
  variables = variables,
  state = "VA",
  year = 2020,
  survey = "acs5",
  geometry = TRUE
)

# View the first few rows of the data
head(acs_data)

```
## Step 4: Rename columns for clarity
```{r}
colnames(acs_data) <- c("GEOID", "NAME", "variable","Median_Income", "moe","geometry")
```

## Step 5: Perform Basic Analysis
Perform basic analysis to explore income levels.

```{r}
# Summary statistics for median income
summary(acs_data$Median_Income)

# Plot median income distribution
ggplot(acs_data, aes(x = Median_Income)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black") +
  labs(title = "Distribution of Median Household Income in Virginia",
       x = "Median Household Income",
       y = "Frequency")
```


## Step 6: Map Visualization
Visualize the data on a map.
```{r}
# Plot median income on a map
ggplot(acs_data) +
  geom_sf(aes(fill = Median_Income)) +
  scale_fill_viridis_c() +
  labs(title = "Median Household Income by Census Tract in Virginia",
       fill = "Median Income")
```
## Step 7: Download ACS/census data for multiple states
```{r}
acs_data <- get_acs(geography = "tract", 
                   variables = "B19013_001",
                   state = c("DC", "MD", "VA"),
                   geometry = TRUE)
colnames(acs_data) <- c("GEOID", "NAME", "variable","Median_Income", "moe","geometry")
```

## Step 7: use mapview to visualize the results:
```{r}
mapview(acs_data,z='Median_Income')
```

## Use the Gini index to quantify inequality in the DMV area
Now that we have income data for DC, VA, and MD, let's focus on the DMV (District of Columbia, Maryland, and Virginia) area for some basic analysis. Specifically, we will quantify inequality using the Gini index. The Gini index is a measure of statistical dispersion intended to represent the income inequality within a region. A Gini index of 0 represents perfect equality, while an index of 1 indicates maximal inequality.

## Step 1: important DMV shapefile
```{r}
dmv<-st_read('dmv.shp')
plot(dmv)
```
## Step 2: Select tracts by location (dmv boundary)
```{r}
# Transform the CRS of the ACS data to match the DMV shapefile
acs_data <- st_transform(acs_data, st_crs(dmv))

# Perform the spatial subset
dmv_data <- acs_data[dmv, ,op = st_intersects]
```

## Step 3: use mapview to visualize the results:
```{r}
mapview(dmv_data,z='Median_Income')
```


## Step 3: Calculate the Gini index for the entire DMV area
### install 'ineq' library before you run the following code
### install.packages('ineq')
```{r}
library(ineq)
gini_index <- Gini(dmv_data$Median_Income)

# Print the Gini index
print(gini_index)

```
## Step 4: Write Data to a Shapefile
Save the processed data as a shapefile.
```{r}
st_write(dmv_data, "acs_data_dmv.shp", delete_layer = TRUE)
```
## Lab questions:
1. Identify and Report the Tract with the Highest Median Income in DMV

2. Identify and Report the Tract with the Lowest Median Income in DMV

3. Calculate the Gini index for the Richmond, VA Metro Area. The spatial boundary of the Richmond Metro Area [Richmond_metro.shp] is included in the Lab 4 folder. You will need to revise the script, particularly the section on selecting by location, as well as parts of the lab steps to obtain the answer.

4. Create a map visualizing the median household income for each tract within the Richmond VA Metro area using the tmap package. Attach the resulting map.

