---
title: "R Programming for Geospatial Applications"
output:
  powerpoint_presentation: default
  ioslides_presentation: default
date: "2024-07-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline

- *Download geospatial data using R libraries *  

<br>

<br>

- *For loop (examples)*  

## Commonly use geospatial data
- Administrative boundaries 
- Census data (population by block groups, income, education level, county, etc.)
- Roads
- DEM
- NLCD
- The Global Historical Climatology Network (GHCN)
- Soil

## Recreation Park Siting Project (Geog4084/5084)
Please download population (block group level), primary and second roads, DEM, soil, and NLCD 2019 data for Loudoun County, VA 

## Automate data downloading 
- *tidycensus*: designed to help R users get Census data that is pre-prepared for exploration within the tidyverse, and optionally spatially with sf.

<br>

- *tigris*: download census TIGER/Line Shapefiles

<br>

- *FedData*: Functions to automate downloading geospatial data available from several federated data sources (mainly sources maintained by the US Federal government). 

## tidycensus
### replace my api key with your CENSUS API key
### .Renviron file so it can be called securely without being stored in your code
```{r,echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidycensus)
census_api_key("e23a9a88f3a3911be51aed1a0e9c595a10e35b59",overwrite = TRUE)
readRenviron("~/.Renviron")
```
## tidycensus
```{r,echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidycensus)
pop2010 <- get_decennial(geography = "county", 
                         variables = 'P001001', 
                         year = 2010, state="VA", 
                         geometry = TRUE)
```

## tidycensus
```{r,echo=TRUE, message=FALSE, warning=FALSE}
head(pop2010)
```

## tidycensus
```{r,echo=TRUE, message=FALSE, warning=FALSE}
plot(pop2010["value"])
```

## Downloading ACS Data Using tidycensus
American Community Survey (ACS): A detailed annual survey conducted by the U.S. Census Bureau.

```{r,echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidycensus)
library(dplyr)
##Download median household income for all counties in 2019.
income_data <- get_acs(geography = "county",
                       variables = "B19013_001",
                       year = 2019,
                       geometry=TRUE,
                       survey = "acs5")
```

## view income data
```{r,echo=TRUE, message=FALSE, warning=FALSE}
head(income_data)
print(dim(income_data))
```


## plot income data
```{r,echo=TRUE, message=FALSE, warning=FALSE}
library(tmap)
library(dplyr)
library(stringr)
ca_income_data <- income_data %>% filter(str_starts(GEOID, "06"))
qtm(ca_income_data,fill='estimate')
```
## tigris

- tigris: An R package that provides access to geographic data from the U.S. Census Bureau.
```{r,echo=TRUE, message=FALSE, warning=FALSE}
library(tigris)
library(sf)
options(tigris_class = "sf")
ca_roads <- primary_secondary_roads(state = "VA", year = 2020)
plot(st_geometry(ca_roads))

```

## FedData library
- The National Elevation Dataset digital elevation models 
- The National Hydrography Dataset (USGS)
- The Soil Survey Geographic (SSURGO) database 
- The Global Historical Climatology Network (GHCN)
- Daymet gridded estimates of daily weather parameters for North 	America, version 3
- The International Tree Ring Data Bank
- National Land Cover Database (NLCD).

## FedData example
```{r,echo=TRUE, message=FALSE, warning=FALSE}
library(FedData)

```

## Outline

- *Download geospatial data using R libraries *  

<br>

<br>

- *For loop (examples)*  

## For loop
- A for loop allows you to iterate over a sequence of elements, performing a set of operations for each element.

```{r,echo=TRUE, message=FALSE, warning=FALSE}
for (i in 1:5) {
  print(i)
}
```
## Another example

```{r,echo=TRUE, message=FALSE, warning=FALSE}
numbers <- c(1, 2, 3, 4, 5)
total <- 0

for (num in numbers) {
  total <- total + num
  print(total)
}

print(total)
```
## GIS task: download ACS income data for all states













